\textbf{COMP0085 Summative Assignment}

### 2022-12-28 submission.tex
update assignment

The free energy is can be calculated as:
\[\mathcal{F}(q, \theta) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]

%Thus,
%\[\mathcal{F}(Q, \theta) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{i=1}^{K} q_{i} (s_i)} + H\left[ \prod_{i=1}^{K} q_{i} (s_i)\right] \]

@@

\[\mathcal{F}(Q, \theta) = \langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} + \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})} + H\left[ q(\textbf{s})\right] \]

@@

\[
\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \langle s_i s_j \rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]

@@

\[
\langle \log P(\textbf{x} | \textbf{s}, \theta)\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]

where $\langle s_i s_i \rangle_{q_{i} (s_i)} = \langle s_i \rangle_{q_{i} (s_i)}$ because $s_i \in \{0, 1\}$.

@@

\[ \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \langle s_i \rangle_{q_{i} (s_i)} \log \pi_i + (1-\langle s_i\rangle_{q_{i} (s_i)}) \log (1-\pi_i)}\]

@@

\[ \langle \log P(\textbf{s} | \theta)\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\]

@@

\[\frac{\partial}{\partial q_i}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_i -1)\right) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i) - 1 + \lambda^{LG}\]

@@

\[\log q_i(s_i) = \langle \log P(\textbf{x}, \textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} - 1 + \lambda^{LG}\]

@@

\[\langle \log P(\textbf{x} |  \textbf{s}, \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} \propto  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \langle s_k \rangle_{\prod_{j\neq i} q_j(s_j)} \mu_i   + \sum_{k=1}^{K} \sum_{j=1}^{K} \langle s_k s_j \rangle_{\prod_{j\neq i} q_j(s_j)} \right)\]

@@

\[ \langle \log P(\textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} = \sum_{k=1}^{K} \langle s_k \rangle_{\prod_{j\neq i} q_j(s_j)} \log \pi_k + (1-\langle s_k\rangle_{\prod_{j\neq i} q_j(s_j)}) \log (1-\pi_k)}\]

@@

\[\log q_i(s_i) \propto  \log P(\textbf{x} |  \textbf{s}, \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} + \langle \log P(\textbf{s} | \theta)\rangle_{\prod_{j\neq i} q_j(s_j)} \]

@@

\[\mu = \left( \langle \textbf{s}\textbf{s}^T \rangle_{q(\textbf{s})} \right)^{-1}\langle \textbf{s}\rangle_{q(\textbf{s})} \textbf{x}\]

@@

@@

>>>>

The free energy is can be calculated as:
\[\mathcal{F}(q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{q(\textbf{s})} + H[Q(\textbf{s})]\]

%Thus,
%\[\mathcal{F}(Q, \theta) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{i=1}^{K} q_{i} (s_i)} + H\left[ \prod_{i=1}^{K} q_{i} (s_i)\right] \]

@@

\[\mathcal{F}(Q, \theta) = \left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} + \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})} + H\left[ q(\textbf{s})\right] \]

@@

\[
\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)   -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \mu_i   + \sum_{i=1}^{K} \sum_{j=1}^{K} \left\langle s_i s_j \right\rangle_{q_{i} (s_i) q_{j} (s_j)} \mu_i^T\mu_j \right)\]

@@

\[
\left\langle \log P(\textbf{x} | \textbf{s}, \theta)\right\rangle_{q(\textbf{s})} \\
= -\frac{d}{2} \log (2 \pi \sigma^2)  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{i=1}^{K}  \lambda_i  \mu_i   + \sum_{i=1}^{K} \sum_{j=1, j\neq i}^{K}  \lambda_i \lambda_j \mu_i^T\mu_j + \sum_{i=1}^{K}  \lambda_i \mu_i^T\mu_i \right)\]

where $\left\langle s_i s_i \right\rangle_{q_{i} (s_i)} = \left\langle s_i \right\rangle_{q_{i} (s_i)}$ because $s_i \in \{0, 1\}$.

@@

\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \left\langle s_i \right\rangle_{q_{i} (s_i)} \log \pi_i + (1-\left\langle s_i\right\rangle_{q_{i} (s_i)}) \log (1-\pi_i)}\]

@@

\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{q(\textbf{s})}= \sum_{i=1}^{K} \lambda_i \log \pi_i + (1-\lambda_i) \log (1-\pi_i)}\]

@@

\[\frac{\partial}{\partial q_i}\left( \mathcal{F}(q, \theta) + \lambda^{LG} \int q_i -1)\right) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - \log q_i(s_i) - 1 + \lambda^{LG}\]

@@

\[\log q_i(s_i) = \left\langle \log P(\textbf{x}, \textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} - 1 + \lambda^{LG}\]

@@

\[\left\langle \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} \propto  -\frac{1}{2 \sigma^2} \left(\textbf{x}^T\textbf{x} - 2 \textbf{x}^T\sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{j\neq i} q_j(s_j)} \mu_i   + \sum_{k=1}^{K} \sum_{j=1}^{K} \left\langle s_k s_j \right\rangle_{\prod_{j\neq i} q_j(s_j)} \right)\]

@@

\[ \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} = \sum_{k=1}^{K} \left\langle s_k \right\rangle_{\prod_{j\neq i} q_j(s_j)} \log \pi_k + (1-\left\langle s_k\right\rangle_{\prod_{j\neq i} q_j(s_j)}) \log (1-\pi_k)}\]

@@

\[\log q_i(s_i) \propto  \log P(\textbf{x} |  \textbf{s}, \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} + \left\langle \log P(\textbf{s} | \theta)\right\rangle_{\prod_{j\neq i} q_j(s_j)} \]

@@

\[\mu = \left( \left\langle \textbf{s}\textbf{s}^T \right\rangle_{q(\textbf{s})} \right)^{-1}\left\langle \textbf{s}\right\rangle_{q(\textbf{s})} \textbf{x}\]

@@

We begin with the log joint:

\[\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = \log P(\textbf{x}|  \textbf{s}, \textbf{A}, \Psi,  \eta)
+ \log P( \textbf{s}| \pi, \eta) + \log P( \pi| \eta) + \log P( \textbf{A}| \eta) + \log P( \Psi| \eta)
\]

where $\eta$ is a collection of all hyperparameters.

We know:

\[P(\textbf{x}|  \textbf{s}, \textbf{A}, \Psi,  \eta) = \frac{1}{(2\pi)^{\frac{D}{2}} |\Psi|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As})\right)\]

\[P( \textbf{s}| \pi, \eta) = \prod_{k=1}^{K} \pi_k^{s_i} (1-\pi_k)^{1-s_k}\]

\[P( \pi| \eta) = \prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)}\]

We choose the conjugate priors:

\[P( \textbf{A}| \eta) = \mathcal{N}(\textbf{A}| \mu_{\textbf{A}}, \Sigma_{\textbf{A}}) =\frac{1}{(2\pi)^{\frac{DK}{2}} |\Sigma_{\textbf{A}}|^{\frac{1}{2}}} \exp\left(-\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\right)\]

\[P( \Psi| \eta) = \prod_{d=1}^D InvGamma(\Psi_{dd}| a_d, b_d) = \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-a_d-1}\exp(-\frac{b_d}{\Psi_{dd}})\]

a Gaussian prior on $\textbf{A}$ and a product of inverse gamma distributions on $\Psi$ where we assume $\Psi$ is a diagonal matrix.

Combining, we have our expression:

\[
\begin{array}{l}
\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = -\frac{D}{2} \log (2 \pi) - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \\
\tab \tab \tab \tab \tab + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\\
\tab \tab \tab \tab \tab + \sum_{i=1}^K (\alpha-1) \log \pi_k + (\beta - 1) \log (1-\pi_k) - \log B(\alpha, \beta)\\
\tab \tab \tab \tab \tab -\frac{DK}{2} \log (2 \pi) - \frac{1}{2} \log(|\Sigma_{\textbf{A}}|)   -\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\\
\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - 1) \log\Psi_{dd} - \frac{b_d}{\Psi_{dd}} - \log \Gamma(a_d)\\
\end{array}
\]

%Simplifying:
%
%\[
%\begin{array}{l}
%\log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) = -\frac{D}{2} \log (2 \pi)  -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \\
%\tab \tab \tab \tab \tab + \sum_{i=1}^K (s_k+\alpha-1) \log \pi_k + (\beta - s_k) \log (1-\pi_k) - \log B(\alpha, \beta)\\
%\tab \tab \tab \tab \tab -\frac{DK}{2} \log (2 \pi) - \frac{1}{2} \log(|\Sigma_{\textbf{A}}|)   -\frac{1}{2}\textbf{A}^T \Sigma_{\textbf{A}}^{-1}\textbf{A}\\
%\tab \tab \tab \tab \tab + \sum_{d=1}^D a_d \log b_d + (-a_d - \frac{3}{2}) \log\Psi_{dd} - \frac{b_d}{\Psi_{dd}} - \log \Gamma(a_d)\\
%\end{array}
%\]

For the Variational Bayes expectation step, we minimise $\textbf{KL}[q_{\texbf{s}}(\textbf{s} | \text{everything else}) \| P(\textbf{s} | \text{everything else})]$ by setting:

\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\theta)}\]

where $\theta$ denotes the parameters $\pi$, $\textbf{A}$, $\Psi$, $\eta$.

\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) + \sum_{k=1}^{K} s_k \log \pi_k + (1-s_k) \log (1-\pi_k)}\right\rangle_{q(\theta)}\]

\[q_{\texbf{s}}(\textbf{s}) \propto \exp \left\langle -\frac{1}{2}\left(\textbf{s}^T\textbf{A}^T\Psi^{-1}\textbf{As} - 2\textbf{s}^T\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right)\right\rangle_{q(\theta)}\]

By inspection, we can see:

\[q_{\texbf{s}}(\textbf{s}) \propto \mathcal{N}(s | \mu_\textbf{s}^*, \Sigma_\textbf{s}^*)\]

where

\[\Sigma_\textbf{s}^* = \left\langle \left(\textbf{A}^T\Psi^{-1}\textbf{A}\right)^{-1}\right\rangle_{q(\theta)}\]

\[\mu_\textbf{s}^* = \left\langle \left(\textbf{A}^T\Psi^{-1}\textbf{A}\right)^{-1}\left(\textbf{A}^T\Psi^{-1} \textbf{x}  + 2\log\frac{\pi}{1-\pi}\right)\right\rangle_{q(\theta)}\]

the E step updates.

For the Variational Bayes maximisation step, we set:

\[q_{\theta}(\theta) \propto P(\theta) \exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q(\textbf{s})}\]

assuming the factorisation:

\[q_{\theta}(\theta) = q_{\pi}(\pi)q_{\Psi}(\Psi)q_{\textbf{A}}(\textbf{A})\]

we can calculate each factor independently.

For $q_{\pi}(\pi)$:

\[q_{\pi}(\pi) \propto P(\pi)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]

Substituting the appropriate terms:

\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{i=1}^K s_k \log \pi_k + (1 - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]

We see:

\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{\alpha  + \left\langle s_k\right\rangle_{q_{s_k}} - 1}(1-\pi_k)^{\beta-\left\langle s_k\right\rangle_{q_{s_k}}}}{B(\alpha, \beta)}\]

\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta(\alpha + \left\langle s_k\right\rangle_{q_{s_k}} , \beta+(1-\left\langle s_k\right\rangle_{q_{s_k}}}))\]

%Substituting the appropriate terms:
%
%\[q_{\pi}(\pi) \propto \left(\prod_{k=1}^{K} \frac{\pi_k^{\alpha - 1}(1-\pi_k)^{\beta-1}}{B(\alpha, \beta)} \right)\exp \left\langle \sum_{i=1}^K (s_k+\alpha-1) \log \pi_k + (\beta - s_k) \log (1-\pi_k) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \pi}(\theta)}\]
%
%We see:
%
%\[q_{\pi}(\pi) \propto \prod_{k=1}^{K} \frac{\pi_k^{2\alpha - 2 + \left\langle s_k\right\rangle_{q_{s_k}}}(1-\pi_k)^{2\beta-\left\langle s_k\right\rangle_{q_{s_k}}}}{B(\alpha, \beta)}\]
%
%\[q_{\pi}(\pi) = \prod_{k=1}^{K}  Beta(2\alpha -2 + \left\langle s_k\right\rangle_{q_{s_k}} , 2\beta+1-\left\langle s_k\right\rangle_{q_{s_k}}})\]

For $q_{\Psi}(\Psi)$:

\[q_{\Psi}(\Psi) \propto P(\Psi)\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi}(\theta)}\]

Substituting the appropriate terms:

\[
q_{\Psi}(\Psi) \propto \left( \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-a_d-1}\exp(-\frac{b_d}{\Psi_{dd}}) \right)\exp \left\langle  - \frac{1}{2} \sum_{d=1}^{D}\log\Psi_{dd}   -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \Psi(\theta)}}
\]

We see:

\[
q_{\Psi}(\Psi) \propto  \prod_{d=1}^D \frac{b_d^{a_d}}{\Gamma(a_d)}\Psi_{dd}^{-(a_d+\frac{1}{2})-1}\exp\left(-\frac{b_d+\frac{1}{2} \left\langle (x_d-\textbf{A}_d\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_d}(\textbf{A}_d)}}{\Psi_{dd}}\right) }
\]

where $\textbf{A}_d \in \mathbb{R}^{1 \times K}$ is the $d^{th}$ row of $\textbf{A}$.

\[
q_{\Psi}(\Psi) = \prod_{d=1}^D  InvGamma\left(\Psi_{dd} \left| a_d+\frac{1}{2}, b_d+\frac{1}{2} \left\langle (x_d-\textbf{A}_d\textbf{s})^2 \right\rangle_{q_{\textbf{s}}(\textbf{s})q_{\textbf{A}_d}(\textbf{A}_d)}\right)
\]

For $q_{\textbf{A}}(\textbf{A})$:

\[q_{\textbf{A}}(\textbf{A})  \propto P(\textbf{A})\exp \left\langle \log P(\textbf{x}, \textbf{s}, \pi, \textbf{A}, \Psi | \eta) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]

\[q_{\textbf{A}}(\textbf{A}) \propto \exp\left(-\frac{1}{2}(\textbf{A} - \mu_{\textbf{A}})^T \Sigma_{\textbf{A}}^{-1}(\textbf{A} - \mu_{\textbf{A}})\right) \exp \left\langle -\frac{1}{2}(\textbf{x}-\textbf{As})^T \Psi^{-1} (\textbf{x}-\textbf{As}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]

\[q_{\textbf{A}}(\textbf{A}) \propto \exp\left(-\frac{1}{2}(\textbf{A}^T \Sigma_{\textbf{A}}^{-1}\textbf{A} -2 \textbf{A}^T \Sigma_{\textbf{A}}^{-1}  \mu_{\textbf{A}})\right) \exp \left\langle -\frac{1}{2}(\textbf{s}^T\textbf{A}^T \Psi^{-1} \textbf{A}\textbf{s} - 2 \textbf{s}^T\textbf{A}^T \Psi^{-1} \textbf{x}) \right\rangle_{q_{\textbf{s}}(\textbf{s}) q_{\neg \textbf{A}}(\theta)}\]

@@

We can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features

Place prior on $\sigma^2$ and optimise with respect to the distributions would cause some to diverge and only relevant latent dimensions will remain.
This gives us a value for $K$, the number of latent factors that haven't diverged.
